---
title: "Spatial.analyses"
author: "Smriti Pehim Limbu"
date: "2026-05-20"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Install packages
```{r}
install.packages("libranian")
library(librarian)
shelf(tidyverse, dplyr, readr, codyn, lme4, lmerTest, ggrepel, tictoc, ggpubr, nlme, brms, cmdstanr, MatchIt, cobalt, spdep, bayesplot, tidybayes, posterior, corrplot, rcompanion, ggplot2, patchwork, scales, nnet, tibble, performance)
install.packages(
  "cmdstanr",
  repos = c("https://mc-stan.org/r-packages/", getOption("repos"))
)
library(cmdstanr)
cmdstanr::check_cmdstan_toolchain(fix = TRUE) 
cmdstanr::install_cmdstan()
cmdstanr::check_cmdstan_toolchain()
```
#data upload
```{r}
df_all_with_slope_aspect <- read.csv("df_all_with_slope_aspect.csv")

df_spatial <- df_all_with_slope_aspect %>%
  mutate(
    Land_use = as.character(Land_use),
    MERGE_SRC = as.character(MERGE_SRC),
    Land_use = case_when(
      Land_use == "Managed Forest" ~ "Historically forested",
      Land_use == "Old growth" ~ "Historically forested",
      Land_use == "Managed/CutRegrowth" ~ "Historically forested",
      Land_use == "Pasture1973" ~ "Historically pasture",
      TRUE ~ Land_use
    ),
    Soil_type = case_when(
      MERGE_SRC == "Glaciofluvial_Dissolve" ~ "Glaciolacustrine",
      MERGE_SRC == "Till_Dissolve" ~ "Till",
      TRUE ~ MERGE_SRC
    ),

    Elevation = Elevation.m.,

    Clusters_id = as.factor(Clusters_id),
    Land_use = as.factor(Land_use),
    Soil_type = as.factor(Soil_type),
    Elevation_sc = as.numeric(scale(Elevation)),
    Slope_sc = as.numeric(scale(Slope)),
    Aspect_sc = as.numeric(scale(Aspect)),
    UTM18N_Easting_sc = as.numeric(scale(UTM18N_Easting)),
    UTM18N_Northing_sc = as.numeric(scale(UTM18N_Northing))
  ) %>%
  drop_na(
    Clusters_id,
    Land_use,
    Soil_type,
    Elevation_sc,
    Slope_sc,
    Aspect_sc,
    UTM18N_Easting_sc,
    UTM18N_Northing_sc
  ) %>%
  mutate(
    Aspect_rad = Aspect * pi / 180,
    northness = cos(Aspect_rad),
    eastness = sin(Aspect_rad),
    northness_sc = as.numeric(scale(northness)),
    eastness_sc = as.numeric(scale(eastness))
  )

# Set reference cluster 1 
df_spatial$Clusters_id <- relevel(df_spatial$Clusters_id, ref = "1")
#mutate spatial block

df_spatial <- df_spatial %>%
  mutate(
    x_bin = cut(UTM18N_Easting_sc, breaks = 10),
    y_bin = cut(UTM18N_Northing_sc, breaks = 10),
    spatial_block = interaction(x_bin, y_bin, drop = TRUE)
  )

```

#Analysis (check for correlation between land use and environment)
```{r}
#Check for correlation for numerical variables
numeric_vars <- df_spatial %>%
  select(
    Elevation,
    Slope,
    northness,
    eastness,
    UTM18N_Easting,
    UTM18N_Northing
  )

cor_mat <- cor(numeric_vars, use = "pairwise.complete.obs")

cor_mat

corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  tl.col = "black",
  tl.srt = 45
)

#for categorical variables

landuse_soil_table <- table(df_spatial$Land_use, df_spatial$Soil_type)

landuse_soil_table

prop.table(landuse_soil_table, margin = 1)

chisq.test(landuse_soil_table)

cramerV(
  landuse_soil_table,
  bias.correct = TRUE
)

```

#Final figures for paper (supplemental env and management summaries & observed vs predicted figure)
```{r}
####supplemental showing the environment and management summaries####

cluster_cols <- c(
  "1" = "#1F77B4",
  "2" = "#FF7F0F",
  "3" = "#2CA02C"
)

df_plot <- df_spatial %>%
  mutate(
    Clusters_id = factor(Clusters_id, levels = c("1", "2", "3"))
  )

p_elev <- ggplot(df_plot, aes(x = Clusters_id, y = Elevation_sc, fill = Clusters_id)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.4) +
  scale_fill_manual(values = cluster_cols) +
  theme_classic() +
  labs(
    x = "Cluster",
    y = "Scaled elevation",
    fill = "Cluster"
  ) +
  theme(legend.position = "none")

p_slope <- ggplot(df_plot, aes(x = Clusters_id, y = Slope_sc, fill = Clusters_id)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.4) +
  scale_fill_manual(values = cluster_cols) +
  theme_classic() +
  labs(
    x = "Cluster",
    y = "Scaled slope",
    fill = "Cluster",
    title = "Slope"
  ) +
  theme(legend.position = "none")

p_north <- ggplot(df_plot, aes(x = Clusters_id, y = northness_sc, fill = Clusters_id)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.4) +
  scale_fill_manual(values = cluster_cols) +
  theme_classic() +
  labs(
    x = "Cluster",
    y = "Scaled northness",
    fill = "Cluster",
    title = "Northness"
  ) +
  theme(legend.position = "none")

p_east <- ggplot(df_plot, aes(x = Clusters_id, y = eastness_sc, fill = Clusters_id)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.4) +
  scale_fill_manual(values = cluster_cols) +
  theme_classic() +
  labs(
    x = "Cluster",
    y = "Scaled eastness",
    fill = "Cluster",
    title = "Eastness"
  ) +
  theme(legend.position = "none")

land_use_summary <- df_plot %>%
  count(Clusters_id, Land_use) %>%
  group_by(Clusters_id) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

p_land <- ggplot(land_use_summary, aes(x = Clusters_id, y = prop, fill = Land_use)) +
  geom_col(color = "white", linewidth = 0.2) +
  theme_classic() +
  labs(
    x = "Cluster",
    y = "Proportion",
    fill = "Land-use history",
    title = "Land-use history"
  )

soil_summary <- df_plot %>%
  count(Clusters_id, Soil_type) %>%
  group_by(Clusters_id) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

p_soil <- ggplot(soil_summary, aes(x = Clusters_id, y = prop, fill = Soil_type)) +
  geom_col(color = "white", linewidth = 0.2) +
  theme_classic() +
  labs(
    x = "Cluster",
    y = "Proportion",
    fill = "Soil type",
    title = "Soil type"
  )

fig_env_mgmt <- (p_land | p_soil) /
  (p_elev | p_slope) /
  (p_north | p_east)

fig_env_mgmt
ggsave(
  filename = "figure_cluster_environment_management_summaries.png",
  plot = fig_env_mgmt,
  width = 10,
  height = 10,
  dpi = 300
)

####figure showing predicted cluster map####

cluster_cols <- c(
  "1" = "#1F77B4",
  "2" = "#FF7F0F",
  "3" = "#2CA02C"
)

pred_probs_block <- fitted(m_block, scale = "response")
pred_mean_block <- pred_probs_block[, "Estimate", ]

pred_block_df <- cbind(
  df_spatial,
  as.data.frame(pred_mean_block)
)

pred_block_df$predicted_cluster <- colnames(pred_mean_block)[
  apply(pred_mean_block, 1, which.max)
]

pred_block_df$predicted_cluster_clean <- gsub(
  "P\\(Y = |\\)",
  "",
  pred_block_df$predicted_cluster
)

pred_block_df <- pred_block_df %>%
  mutate(
    predicted_cluster_clean = factor(
      predicted_cluster_clean,
      levels = c("1", "2", "3")
    )
  )

p_predicted_cluster <- ggplot(pred_block_df, aes(
  x = UTM18N_Easting_sc,
  y = UTM18N_Northing_sc,
  color = predicted_cluster_clean
)) +
  geom_point(size = 1) +
  coord_equal() +
  scale_color_manual(values = cluster_cols) +
  theme_classic() +
  labs(
    x = "Scaled UTM Easting",
    y = "Scaled UTM Northing",
    color = "Predicted cluster",
    title = "Predicted cluster membership"
  )

p_predicted_cluster

ggsave(
  filename = "figure_predicted_cluster_map_spatial_block.png",
  plot = p_predicted_cluster,
  width = 7,
  height = 6,
  dpi = 300
)
p_observed_cluster <- ggplot(pred_block_df, aes(
  x = UTM18N_Easting_sc,
  y = UTM18N_Northing_sc,
  color = factor(Clusters_id, levels = c("1", "2", "3"))
)) +
  geom_point(size = 1) +
  coord_equal() +
  scale_color_manual(values = cluster_cols) +
  theme_classic() +
  labs(
    x = "Scaled UTM Easting",
    y = "Scaled UTM Northing",
    color = "Observed cluster",
    title = "Observed cluster membership"
  )

library(patchwork)


p_observed_cluster <- p_observed_cluster +
  labs(color = "Cluster")

p_predicted_cluster <- p_predicted_cluster +
  labs(color = "Cluster")

fig_observed_predicted <- (p_observed_cluster | p_predicted_cluster) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "right",
    plot.tag = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 16, face = "plain"),
    axis.text = element_text(size = 16),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "plain"),
    legend.text = element_text(size = 16)
  )

fig_observed_predicted

ggsave(
  filename = "figure_observed_predicted_cluster_membership.png",
  plot = fig_observed_predicted,
  width = 11,
  height = 5.5,
  dpi = 300
)

```
####Key predictors conditional effect####
```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(brms)

cluster_cols <- c(
  "1" = "#1F77B4",
  "2" = "#FF7F0F",
  "3" = "#2CA02C"
)

ce_land <- conditional_effects(
  m_block,
  effects = "Land_use",
  categorical = TRUE
)

ce_soil <- conditional_effects(
  m_block,
  effects = "Soil_type",
  categorical = TRUE
)

ce_elev <- conditional_effects(
  m_block,
  effects = "Elevation_sc",
  categorical = TRUE
)

ce_slope <- conditional_effects(
  m_block,
  effects = "Slope_sc",
  categorical = TRUE
)

ce_north <- conditional_effects(
  m_block,
  effects = "northness_sc",
  categorical = TRUE
)

ce_east <- conditional_effects(
  m_block,
  effects = "eastness_sc",
  categorical = TRUE
)

clean_ce <- function(ce_object) {
  
  ce_df <- as.data.frame(ce_object[[1]])
  
  # Find the column that stores cluster/category identity
  category_col <- intersect(
    c("category", "cats__", "cat", "resp"),
    names(ce_df)
  )[1]
  
  if (is.na(category_col)) {
    stop(
      "Could not find a category column. Available columns are: ",
      paste(names(ce_df), collapse = ", ")
    )
  }
  
  ce_df %>%
    mutate(
      category_clean = .data[[category_col]],
      category_clean = gsub("P\\(Y = |\\)", "", category_clean),
      category_clean = gsub("^mu", "", category_clean),
      category_clean = factor(category_clean, levels = c("1", "2", "3"))
    )
}

ce_land_df  <- clean_ce(ce_land)
ce_soil_df  <- clean_ce(ce_soil)
ce_elev_df  <- clean_ce(ce_elev)
ce_slope_df <- clean_ce(ce_slope)
ce_north_df <- clean_ce(ce_north)
ce_east_df  <- clean_ce(ce_east)
names(ce_land_df)
head(ce_land_df)

theme_ce <- theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    axis.text.x = element_text(size = 10),
    strip.text = element_text(face = "bold")
  )

##management plot
library(dplyr)
library(ggplot2)
library(patchwork)

cluster_cols <- c(
  "1" = "#1F77B4",
  "2" = "#FF7F0F",
  "3" = "#2CA02C"
)

# Clean labels if needed
ce_land_df <- ce_land_df %>%
  mutate(
    Land_use_clean = case_when(
      Land_use == "Historically forested" ~ "Historically\nforested",
      Land_use == "Historically pasture" ~ "Historically\npasture",
      TRUE ~ as.character(Land_use)
    )
  )

ce_soil_df <- ce_soil_df %>%
  mutate(
    Soil_type_clean = as.character(Soil_type)
  )

theme_ce <- theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_ce_land <- ggplot(
  ce_land_df,
  aes(
    x = Land_use_clean,
    y = estimate__,
    color = category_clean,
    group = category_clean
  )
) +
  geom_errorbar(
    aes(ymin = lower__, ymax = upper__),
    width = 0.08,
    linewidth = 0.7,
    position = position_dodge(width = 0.45)
  ) +
  geom_point(
    size = 3,
    position = position_dodge(width = 0.45)
  ) +
  scale_color_manual(values = cluster_cols, name = "Cluster") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(x) paste0(round(x * 100), "%"),
    expand = expansion(mult = c(0, 0.03))
  ) +
  labs(
    x = "Management history",
    y = "Predicted probability",
  ) +
  theme_ce

#soil plot
p_ce_soil <- ggplot(
  ce_soil_df,
  aes(
    x = Soil_type_clean,
    y = estimate__,
    color = category_clean,
    group = category_clean
  )
) +
  geom_errorbar(
    aes(ymin = lower__, ymax = upper__),
    width = 0.08,
    linewidth = 0.7,
    position = position_dodge(width = 0.45)
  ) +
  geom_point(
    size = 3,
    position = position_dodge(width = 0.45)
  ) +
  scale_color_manual(values = cluster_cols, name = "Cluster") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(x) paste0(round(x * 100), "%"),
    expand = expansion(mult = c(0, 0.03))
  ) +
  labs(
    x = "Soil type",
    y = "Predicted probability"
  ) +
  theme_ce

##elevation plot
p_ce_elev <- ggplot(ce_elev_df, aes(
  x = Elevation_sc,
  y = estimate__,
  color = category_clean,
  fill = category_clean
)) +
  geom_ribbon(
    aes(ymin = lower__, ymax = upper__),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = cluster_cols, drop = FALSE) +
  scale_fill_manual(values = cluster_cols, drop = FALSE) + theme_ce +
  labs(
    x = "Scaled elevation",
    y = "Predicted probability",
    color = "Cluster",
    fill = "Cluster"
  )

#slope plot
p_ce_slope <- ggplot(ce_slope_df, aes(
  x = Slope_sc,
  y = estimate__,
  color = category_clean,
  fill = category_clean
)) +
  geom_ribbon(
    aes(ymin = lower__, ymax = upper__),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = cluster_cols, drop = FALSE) +
  scale_fill_manual(values = cluster_cols, drop = FALSE) +
  theme_ce +
  labs(
    x = "Scaled slope",
    y = "Predicted probability",
    color = "Cluster",
    fill = "Cluster"
  )
fig_conditional_effects_main <- (p_ce_land | p_ce_soil) /
  (p_ce_elev | p_ce_slope) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    text = element_text(size = 16),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14)
  )

fig_conditional_effects_main

ggsave(
  filename = "figure_conditional_effects_main.png",
  plot = fig_conditional_effects_main,
  width = 11,
  height = 8.5,
  dpi = 300
)



# Land-use predicted probabilities
land_stats <- ce_land_df %>%
  mutate(
    lower_percent = lower__ * 100,
    estimate_percent = estimate__ * 100,
    upper_percent = upper__ * 100
  ) %>%
  select(
    Land_use,
    cluster = category_clean,
    estimate_percent,
    lower_percent,
    upper_percent
  )

land_stats

soil_stats <- ce_soil_df %>%
  mutate(
    lower_percent = lower__ * 100,
    estimate_percent = estimate__ * 100,
    upper_percent = upper__ * 100
  ) %>%
  select(
    Soil_type,
    cluster = category_clean,
    estimate_percent,
    lower_percent,
    upper_percent
  )

soil_stats

land_change <- ce_land_df %>%
  select(Land_use, category_clean, estimate__) %>%
  tidyr::pivot_wider(
    names_from = Land_use,
    values_from = estimate__
  ) %>%
  mutate(
    change_pasture_minus_forested =
      `Historically pasture` - `Historically forested`
  )

land_change

soil_change <- ce_soil_df %>%
  select(Soil_type, category_clean, estimate__) %>%
  tidyr::pivot_wider(
    names_from = Soil_type,
    values_from = estimate__
  ) %>%
  mutate(
    change_till_minus_glaciolacustrine =
      Till - Glaciolacustrine
  )

soil_change

confusion_table <- table(
  observed = pred_block_df$Clusters_id,
  predicted = pred_block_df$predicted_cluster_clean
)

confusion_table

overall_accuracy <- mean(
  as.character(pred_block_df$Clusters_id) ==
    pred_block_df$predicted_cluster_clean
)

overall_accuracy

library(dplyr)

# Get predicted probabilities from the spatial-block model
pred_probs_block <- fitted(m_block, scale = "response")
pred_mean_block <- pred_probs_block[, "Estimate", ]

# Create prediction dataframe
pred_block_df <- cbind(
  df_spatial,
  as.data.frame(pred_mean_block)
)

# Most likely predicted cluster
pred_block_df$predicted_cluster <- colnames(pred_mean_block)[
  apply(pred_mean_block, 1, which.max)
]

# Clean predicted cluster labels
pred_block_df$predicted_cluster_clean <- gsub(
  "P\\(Y = |\\)",
  "",
  pred_block_df$predicted_cluster
)

# Maximum predicted probability = model confidence
pred_block_df$max_pred_prob <- apply(pred_mean_block, 1, max)

# Correct / incorrect prediction
pred_block_df <- pred_block_df %>%
  mutate(
    Clusters_id = as.character(Clusters_id),
    predicted_cluster_clean = as.character(predicted_cluster_clean),
    correct_prediction = Clusters_id == predicted_cluster_clean
  )

overall_accuracy <- mean(pred_block_df$correct_prediction, na.rm = TRUE)

overall_accuracy

accuracy_by_cluster <- pred_block_df %>%
  group_by(Clusters_id) %>%
  summarise(
    n = n(),
    correct = sum(correct_prediction, na.rm = TRUE),
    accuracy = mean(correct_prediction, na.rm = TRUE),
    mean_confidence = mean(max_pred_prob, na.rm = TRUE),
    median_confidence = median(max_pred_prob, na.rm = TRUE),
    .groups = "drop"
  )

accuracy_by_cluster

accuracy_by_cluster_report <- accuracy_by_cluster %>%
  mutate(
    accuracy_percent = round(accuracy * 100, 1),
    mean_confidence_percent = round(mean_confidence * 100, 1),
    median_confidence_percent = round(median_confidence * 100, 1)
  ) %>%
  select(
    Cluster = Clusters_id,
    n,
    correct,
    accuracy_percent,
    mean_confidence_percent,
    median_confidence_percent
  )

accuracy_by_cluster_report


```
##Final figures for paper (management confounding with environment)
```{r}
colnames(df_spatial)
df_conf <- df_spatial %>%
  mutate(
    Clusters_id = factor(Clusters_id, levels = c("1", "2", "3")),
    Land_use = factor(Land_use),
    Soil_type = factor(Soil_type),
    spatial_block = factor(spatial_block)
  )

# Check factor levels and reference categories
levels(df_conf$Clusters_id)
levels(df_conf$Land_use)
levels(df_conf$Soil_type)

# Cluster 1 is the response reference category
df_conf$Clusters_id <- relevel(df_conf$Clusters_id, ref = "1")


#management and soil type

land_soil_tab <- table(
  Land_use = df_conf$Land_use,
  Soil_type = df_conf$Soil_type
)

# Soil-type proportions within each land-use category
round(prop.table(land_soil_tab, margin = 1), 3)

# Chi-square test
chisq_land_soil <- chisq.test(
  land_soil_tab,
  correct = FALSE
)

chisq_land_soil

# Cramer's V function
cramers_v <- function(tab) {
  chi_test <- suppressWarnings(
    chisq.test(tab, correct = FALSE)
  )

  n_total <- sum(tab)
  min_dimension <- min(
    nrow(tab) - 1,
    ncol(tab) - 1
  )

  sqrt(
    as.numeric(chi_test$statistic) /
      (n_total * min_dimension)
  )
}

cramers_v_land_soil <- cramers_v(land_soil_tab)

cramers_v_land_soil


# Plot soil-type composition within management histories
land_soil_plot_df <- df_conf %>%
  count(Land_use, Soil_type, name = "n") %>%
  group_by(Land_use) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

p_land_soil <- ggplot(
  land_soil_plot_df,
  aes(
    x = Land_use,
    y = proportion,
    fill = Soil_type
  )
) +
  geom_col(
    width = 0.7,
    color = "white",
    linewidth = 0.3
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  theme_classic(base_size = 14) +
  labs(
    x = "Management history",
    y = "Proportion",
    fill = "Soil type"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 25,
      hjust = 1
    )
  )

p_land_soil


#management with other environmental variables

management_environment_summary <- df_conf %>%
  group_by(Land_use) %>%
  summarise(
    n = n(),

    elevation_mean = mean(Elevation_sc, na.rm = TRUE),
    elevation_sd = sd(Elevation_sc, na.rm = TRUE),

    slope_mean = mean(Slope_sc, na.rm = TRUE),
    slope_sd = sd(Slope_sc, na.rm = TRUE),

    northness_mean = mean(northness_sc, na.rm = TRUE),
    northness_sd = sd(northness_sc, na.rm = TRUE),

    eastness_mean = mean(eastness_sc, na.rm = TRUE),
    eastness_sd = sd(eastness_sc, na.rm = TRUE),

    .groups = "drop"
  )

management_environment_summary


# Function to calculate standardized mean differences
calculate_smd <- function(data, outcome, group) {

  temporary_data <- data %>%
    select(
      outcome_value = all_of(outcome),
      group_value = all_of(group)
    ) %>%
    filter(
      !is.na(outcome_value),
      !is.na(group_value)
    ) %>%
    mutate(
      group_value = factor(group_value)
    )

  group_levels <- levels(temporary_data$group_value)

  if (length(group_levels) != 2) {
    stop("Land_use must contain exactly two levels.")
  }

  group_1 <- temporary_data$outcome_value[
    temporary_data$group_value == group_levels[1]
  ]

  group_2 <- temporary_data$outcome_value[
    temporary_data$group_value == group_levels[2]
  ]

  pooled_sd <- sqrt(
    (
      (length(group_1) - 1) * var(group_1) +
        (length(group_2) - 1) * var(group_2)
    ) /
      (
        length(group_1) +
          length(group_2) -
          2
      )
  )

  tibble(
    predictor = outcome,
    reference_group = group_levels[1],
    comparison_group = group_levels[2],
    mean_reference = mean(group_1),
    mean_comparison = mean(group_2),
    standardized_mean_difference =
      (mean(group_2) - mean(group_1)) / pooled_sd
  )
}

management_smd <- bind_rows(
  calculate_smd(
    df_conf,
    "Elevation_sc",
    "Land_use"
  ),
  calculate_smd(
    df_conf,
    "Slope_sc",
    "Land_use"
  ),
  calculate_smd(
    df_conf,
    "northness_sc",
    "Land_use"
  ),
  calculate_smd(
    df_conf,
    "eastness_sc",
    "Land_use"
  )
)

management_smd


# Plot environmental overlap by management history
environment_long <- df_conf %>%
  select(
    Land_use,
    Elevation_sc,
    Slope_sc,
    northness_sc,
    eastness_sc
  ) %>%
  pivot_longer(
    cols = -Land_use,
    names_to = "Predictor",
    values_to = "Value"
  ) %>%
  mutate(
    Predictor = recode(
      Predictor,
      Elevation_sc = "Elevation",
      Slope_sc = "Slope",
      northness_sc = "Northness",
      eastness_sc = "Eastness"
    )
  )

p_management_environment <- ggplot(
  environment_long,
  aes(
    x = Land_use,
    y = Value,
    fill = Land_use
  )
) +
  geom_boxplot(
    width = 0.65,
    alpha = 0.8,
    outlier.size = 0.3
  ) +
  facet_wrap(
    ~ Predictor,
    scales = "free_y",
    nrow = 2
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 25,
      hjust = 1
    )
  ) +
  labs(
    x = "Management history",
    y = "Scaled predictor value"
  )

p_management_environment
```
#Final figures for paper (Bayesian spatial model )
# MANAGEMENT ONLY, ENVIRONMENT ONLY, AND COMBINED
```{r}
priors_cat <- c(
  prior(
    normal(0, 1),
    class = "b",
    dpar = "mu2"
  ),
  prior(
    normal(0, 1),
    class = "b",
    dpar = "mu3"
  ),
  prior(
    normal(0, 2),
    class = "Intercept",
    dpar = "mu2"
  ),
  prior(
    normal(0, 2),
    class = "Intercept",
    dpar = "mu3"
  )
)

#management only model 
m_management <- brm(
  formula =
    Clusters_id ~
    Land_use +
    (1 | spatial_block),

  data = df_conf,
  family = categorical(),
  prior = priors_cat,

  chains = 4,
  cores = 4,
  iter = 1000,
  warmup = 500,

  seed = 123,
  backend = "cmdstanr",

  control = list(
    adapt_delta = 0.95,
    max_treedepth = 12
  )
)

#environment only model 

m_environment <- brm(
  formula =
    Clusters_id ~
    Soil_type +
    Elevation_sc +
    Slope_sc +
    northness_sc +
    eastness_sc +
    (1 | spatial_block),

  data = df_conf,
  family = categorical(),
  prior = priors_cat,

  chains = 4,
  cores = 4,
  iter = 1000,
  warmup = 500,

  seed = 123,
  backend = "cmdstanr",

  control = list(
    adapt_delta = 0.95,
    max_treedepth = 12
  )
)

#combined model 
m_combined <- brm(
  formula = Clusters_id ~
    Land_use +
    Soil_type +
    Elevation_sc +
    Slope_sc +
    northness_sc +
    eastness_sc +
    (1 | spatial_block),
  data = df_spatial,
  family = categorical(),
  prior = priors_cat,
  chains = 4,
  cores = 4,
  iter = 1000,
  warmup = 500,
  seed = 123,
  backend = "cmdstanr",
  control = list(
    adapt_delta = 0.95,
    max_treedepth = 12
  )
)

summary(m_management)
summary(m_environment)
summary(m_combined)


# Compare all three models
loo_all_models <- loo_compare(
  loo_combined,
  loo_environment,
  loo_management
)

loo_all_models

# Does adding management improve the environmental spatial model?
loo_management_added <- loo_compare(
  loo_combined,
  loo_environment
)

loo_management_added


# Compare combined model with management-only model
loo_environment_added <- loo_compare(
  loo_combined,
  loo_management
)

loo_environment_added


# COMPARE LAND-USE COEFFICIENTS BEFORE AND AFTER
# ENVIRONMENTAL ADJUSTMENT

management_only_effects <- as.data.frame(
  fixef(m_management),
  row.names = TRUE
) %>%
  rownames_to_column("parameter") %>%
  filter(
    grepl("Land_use", parameter)
  ) %>%
  mutate(
    model = "Management only"
  )

combined_management_effects <- as.data.frame(
  fixef(m_block),
  row.names = TRUE
) %>%
  rownames_to_column("parameter") %>%
  filter(
    grepl("Land_use", parameter)
  ) %>%
  mutate(
    model = "Management + environment"
  )

management_effect_comparison <- bind_rows(
  management_only_effects,
  combined_management_effects
) %>%
  select(
    model,
    parameter,
    Estimate,
    Est.Error,
    Q2.5,
    Q97.5
  ) %>%
  rename(
    estimate = Estimate,
    posterior_sd = Est.Error,
    lower_95_CrI = Q2.5,
    upper_95_CrI = Q97.5
  )

management_effect_comparison

##diagnostics
write.csv(
  management_environment_summary,
  "management_environment_summary.csv",
  row.names = FALSE
)

write.csv(
  management_smd,
  "management_environment_standardized_differences.csv",
  row.names = FALSE
)

write.csv(
  as.data.frame(continuous_correlation_matrix),
  "continuous_predictor_correlations.csv",
  row.names = TRUE
)

####table for contrast from m_combined model

model_effect_table <- fixef(
  m_combined,
  probs = c(0.025, 0.975)
) %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  
  # Exclude model intercepts
  filter(!str_detect(parameter, "Intercept")) %>%
  
  mutate(
    # Identify the cluster contrast
    contrast = case_when(
      str_detect(parameter, "^mu2_") ~ "Cluster 2 vs Cluster 1",
      str_detect(parameter, "^mu3_") ~ "Cluster 3 vs Cluster 1",
      TRUE ~ NA_character_
    ),

    # Remove mu2_ or mu3_ from parameter name
    predictor_raw = str_remove(parameter, "^mu[23]_"),

    # Create readable predictor names
    predictor = case_when(
      str_detect(
        predictor_raw,
        regex("Land_use.*pasture", ignore_case = TRUE)
      ) ~ "Historically pastured",

      str_detect(
        predictor_raw,
        regex("Soil_type.*Till", ignore_case = TRUE)
      ) ~ "Till soil",

      predictor_raw == "Elevation_sc" ~ "Elevation",
      predictor_raw == "Slope_sc" ~ "Slope",
      predictor_raw == "northness_sc" ~ "Northness",
      predictor_raw == "eastness_sc" ~ "Eastness",

      TRUE ~ predictor_raw
    )
  ) %>%

  transmute(
    Contrast = contrast,
    Predictor = predictor,
    Estimate = Estimate,
    `Posterior SD` = Est.Error,
    `Lower 95% CrI` = Q2.5,
    `Upper 95% CrI` = Q97.5
  ) %>%

  mutate(
    across(
      c(
        Estimate,
        `Posterior SD`,
        `Lower 95% CrI`,
        `Upper 95% CrI`
      ),
      ~ round(.x, 2)
    )
  )

model_effect_table
write.csv(
  model_effect_table,
  "combined_model_fixed_effects.csv",
  row.names = FALSE
)


```
##Final figures for the paper- spatial model conditional effects
```{r}

cluster_cols <- c(
  "1" = "#1F77B4",
  "2" = "#FF7F0E",
  "3" = "#2CA02C"
)

ce_land <- conditional_effects(
  m_combined,
  effects = "Land_use",
  categorical = TRUE,
  re_formula = NA,
  probs = c(0.025, 0.975)
)

ce_soil <- conditional_effects(
  m_combined,
  effects = "Soil_type",
  categorical = TRUE,
  re_formula = NA,
  probs = c(0.025, 0.975)
)

ce_elevation <- conditional_effects(
  m_combined,
  effects = "Elevation_sc",
  categorical = TRUE,
  re_formula = NA,
  probs = c(0.025, 0.975),
  resolution = 100
)

ce_slope <- conditional_effects(
  m_combined,
  effects = "Slope_sc",
  categorical = TRUE,
  re_formula = NA,
  probs = c(0.025, 0.975),
  resolution = 100
)

ce_northness <- conditional_effects(
  m_combined,
  effects = "northness_sc",
  categorical = TRUE,
  re_formula = NA,
  probs = c(0.025, 0.975),
  resolution = 100
)

ce_eastness <- conditional_effects(
  m_combined,
  effects = "eastness_sc",
  categorical = TRUE,
  re_formula = NA,
  probs = c(0.025, 0.975),
  resolution = 100
)

#conditional effects to data frame
clean_ce <- function(ce_object) {

  ce_df <- as.data.frame(ce_object[[1]])

  category_col <- intersect(
    c("cats__", "category", "cat", "resp"),
    names(ce_df)
  )[1]

  if (is.na(category_col)) {
    stop(
      "Could not identify the cluster category column. Columns are: ",
      paste(names(ce_df), collapse = ", ")
    )
  }

  ce_df %>%
    mutate(
      cluster = as.character(.data[[category_col]]),
      cluster = str_remove(cluster, "^P\\(Y = "),
      cluster = str_remove(cluster, "\\)$"),
      cluster = str_remove(cluster, "^mu"),
      cluster = factor(cluster, levels = c("1", "2", "3"))
    )
}

ce_land_df <- clean_ce(ce_land)
ce_soil_df <- clean_ce(ce_soil)
ce_elevation_df <- clean_ce(ce_elevation)
ce_slope_df <- clean_ce(ce_slope)
ce_northness_df <- clean_ce(ce_northness)
ce_eastness_df <- clean_ce(ce_eastness)


ce_land_df <- ce_land_df %>%
  mutate(
    Land_use_label = case_when(
      str_detect(
        as.character(Land_use),
        regex("forest", ignore_case = TRUE)
      ) ~ "Historically\nforested",

      str_detect(
        as.character(Land_use),
        regex("pasture", ignore_case = TRUE)
      ) ~ "Historically\npastured",

      TRUE ~ str_wrap(as.character(Land_use), width = 15)
    )
  )

ce_soil_df <- ce_soil_df %>%
  mutate(
    Soil_type_label = str_replace_all(
      as.character(Soil_type),
      "_",
      " "
    )
  )

theme_effects <- theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 18
    ),
    plot.tag = element_text(
      face = "bold",
      size = 18
    ),
    axis.title = element_text(
      face = "bold",
      size = 18
    ),
    axis.text = element_text(size = 12),
    legend.title = element_text(
      face = "bold",
      size = 18
    ),
    legend.text = element_text(size = 12)
  )

plot_categorical_effect <- function(
    data,
    x_variable,
    x_label,
    panel_title
) {

  dodge <- position_dodge(width = 0.45)

  ggplot(
    data,
    aes(
      x = .data[[x_variable]],
      y = estimate__,
      color = cluster,
      group = cluster
    )
  ) +
    geom_errorbar(
      aes(
        ymin = lower__,
        ymax = upper__
      ),
      width = 0.08,
      linewidth = 0.8,
      position = dodge
    ) +
    geom_point(
      size = 3.2,
      position = dodge
    ) +
    scale_color_manual(
      values = cluster_cols,
      name = "Cluster"
    ) +
    scale_y_continuous(
      labels = function(x) paste0(round(x * 100), "%"),
      breaks = seq(0, 1, by = 0.25),
      expand = expansion(mult = c(0.01, 0.03))
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      x = x_label,
      y = "Predicted probability",
      title = panel_title
    ) +
    theme_effects
}


plot_continuous_effect <- function(
    data,
    x_variable,
    x_label,
    panel_title
) {

  ggplot(
    data,
    aes(
      x = .data[[x_variable]],
      y = estimate__,
      color = cluster,
      fill = cluster
    )
  ) +
    geom_ribbon(
      aes(
        ymin = lower__,
        ymax = upper__
      ),
      alpha = 0.15,
      color = NA
    ) +
    geom_line(linewidth = 1.1) +
    scale_color_manual(
      values = cluster_cols,
      name = "Cluster"
    ) +
    scale_fill_manual(
      values = cluster_cols,
      guide = "none"
    ) +
    scale_y_continuous(
      labels = function(x) paste0(round(x * 100), "%"),
      breaks = seq(0, 1, by = 0.25),
      expand = expansion(mult = c(0.01, 0.03))
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      x = x_label,
      y = "Predicted probability",
      title = panel_title
    ) +
    theme_effects
}


#plot
p_land <- plot_categorical_effect(
  data = ce_land_df,
  x_variable = "Land_use_label",
  x_label = "Management history",
  panel_title = "Management history"
)

p_soil <- plot_categorical_effect(
  data = ce_soil_df,
  x_variable = "Soil_type_label",
  x_label = "Soil type",
  panel_title = "Soil type"
)

p_elevation <- plot_continuous_effect(
  data = ce_elevation_df,
  x_variable = "Elevation_sc",
  x_label = "Scaled elevation",
  panel_title = "Elevation"
)

p_slope <- plot_continuous_effect(
  data = ce_slope_df,
  x_variable = "Slope_sc",
  x_label = "Scaled slope",
  panel_title = "Slope"
)

p_northness <- plot_continuous_effect(
  data = ce_northness_df,
  x_variable = "northness_sc",
  x_label = "Scaled northness",
  panel_title = "Northness"
)

p_eastness <- plot_continuous_effect(
  data = ce_eastness_df,
  x_variable = "eastness_sc",
  x_label = "Scaled eastness",
  panel_title = "Eastness"
)

#combine figures
fig_all_effects <- (
  p_land | p_soil
) / (
  p_elevation | p_slope
) / (
  p_northness | p_eastness
) +
  plot_layout(
    guides = "collect"
  ) +
  plot_annotation(
    tag_levels = "A"
  ) &
  theme(
    legend.position = "bottom"
  )

fig_all_effects

ggsave(
  filename = "figure_all_model_conditional_effects.png",
  plot = fig_all_effects,
  width = 11,
  height = 13,
  dpi = 300
)


```
##Model accuracy
```{r}

# posterior draws × observations × clusters
epred <- posterior_epred(
  m_combined,
  re_formula = NULL
)

dim(epred)


# Result: observations × clusters
mean_probabilities <- apply(
  epred,
  MARGIN = c(2, 3),
  FUN = mean
)

# Obtain cluster labels from the model output
cluster_levels <- dimnames(epred)[[3]]

if (is.null(cluster_levels)) {
  cluster_levels <- levels(
    factor(m_combined$data$Clusters_id)
  )
}

colnames(mean_probabilities) <- cluster_levels


# Assign each stem to its most probable cluster

predicted_cluster <- cluster_levels[
  max.col(
    mean_probabilities,
    ties.method = "first"
  )
]

observed_cluster <- as.character(
  m_combined$data$Clusters_id
)

# Check that model predictions and observations align
stopifnot(
  length(predicted_cluster) == length(observed_cluster)
)

#overall accuracy
overall_accuracy <- mean(
  predicted_cluster == observed_cluster
)

cat(
  "Overall in-sample accuracy:",
  round(overall_accuracy * 100, 1),
  "%\n"
)

#accuracy by cluster
accuracy_by_cluster <- tibble(
  observed = observed_cluster,
  predicted = predicted_cluster
) %>%
  group_by(observed) %>%
  summarise(
    n_stems = n(),
    n_correct = sum(predicted == observed),
    accuracy = mean(predicted == observed),
    .groups = "drop"
  ) %>%
  mutate(
    accuracy_percent = round(accuracy * 100, 1),
    cluster = paste("Cluster", observed)
  ) %>%
  select(
    cluster,
    n_stems,
    n_correct,
    accuracy_percent
  )

accuracy_by_cluster



```
